---
title: "AlbumData"
author: "Daniel DeFoe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning  

```{r}
library(tidyverse)

source("frostFunctions.R")
```


```{r}
songAttrsDf = read.csv("Data/songAttributes.csv")
riaaDf = read.csv("Data/RIAACertifications.csv")
grammyDf = read.csv("Data/grammyWinners.csv")
artistDf = read.csv("Data/ArtistInfo.csv")
billboardDf = read.csv("Data/billboardRankings.csv")
albumDatesGenres = read.csv("Data/albReleaseDatesGenres.csv")
```



```{r}
colnames(billboardDf)[colnames(billboardDf) == "Weekly.rank"] = "BillboardWeekRank"
colnames(billboardDf)[colnames(billboardDf) == "Peak.position"] = "PeakPosBillboard"
colnames(billboardDf)[colnames(billboardDf) == "Weeks.on.chart"] = "WeeksOnBillboard"
colnames(billboardDf)[colnames(billboardDf) == "Date"] = "ReleaseDate"
colnames(billboardDf)[colnames(billboardDf) == "Writing.Credits"] = "WritingCredits"


colnames(riaaDf)[colnames(riaaDf) == "Status"] = "RiaaStatus"
colnames(riaaDf)[colnames(riaaDf) == "Title"] = "Name"


colnames(grammyDf)[colnames(grammyDf) == "Award"] = "GrammyAward"
colnames(grammyDf)[colnames(grammyDf) == "SongTitle"] = "Name"
colnames(grammyDf)[colnames(grammyDf) == "Genre"] = "GrammyGenre"


colnames(songAttrsDf)[colnames(songAttrsDf) == "Track"] = "Name"
```




```{r}
artistDf["Genres"] = lapply(artistDf["Genres"], stringColStandardizer)
artistDf$Genres = tolower(artistDf$Genres)

billboardDf["Artists"] = lapply(billboardDf["Artists"], artistPunctuationStrip)
billboardDf[c("Artists","Name","Genre","WritingCredits", "Features")] = lapply(billboardDf[c("Artists","Name","Genre","WritingCredits", "Features")], stringColStandardizer) #not lyrics
billboardDf$ReleaseDate = as.Date(billboardDf$ReleaseDate, format = "%B %d, %Y")
billboardDf$Week = as.character(billboardDf$Week)
billboardDf$Genre = tolower(billboardDf$Genre)

grammyDf["Artist"] = lapply(grammyDf["Artist"], artistPunctuationStrip)
grammyDf[c("GrammyGenre", "Album", "Artist")] = lapply(grammyDf[c("GrammyGenre", "Album",  "Artist")],stringColStandardizer)
grammyDf$GrammyGenre = tolower(grammyDf$GrammyGenre)

songAttrsDf["Artist"] = lapply(songAttrsDf["Artist"], artistPunctuationStrip)
songAttrsDf[c("Album", "Artist", "Name")] = lapply(songAttrsDf[c("Album", "Artist", "Name")], stringColStandardizer)
songAttrsDf["Explicit"] = lapply(songAttrsDf["Explicit"], as.logical)
songAttrsDf["Mode"] = lapply(songAttrsDf["Mode"], as.factor)

riaaDf["Artist"] = lapply(riaaDf["Artist"], artistPunctuationStrip)
riaaDf[c("Album","Artist","Label")]=lapply(riaaDf[c("Album","Artist","Label")],stringColStandardizer)
riaaDf[c("Album")]=lapply(riaaDf[c("Album")],riaaSuffixStrip)

albumDates["ReleaseDate"] = lapply(albumDates["ReleaseDate"], as.Date)

```





## Engineer Features
Get the data in the form it is needed, by extracting features of each artist's albums individually based on :
  - song qualities
  - previous album qualities
  - album's place in the greater music industry scope
  
First the presence of genre tags in the Hot100 by year will determine the most popular genres per year. 
```{r}
yearlyPopularGenre = popularGenreGetter(billboardDf)
print(yearlyPopularGenre)
```

Step-by-step the datasets will be grouped together

```{r}
riaaDf = riaaDf %>%
  select(-X)
```

```{r}
albumYear = grammyDf %>%
  select(Artist, Album, GrammyYear) %>% 
  distinct()
```

```{r}
grammyDf = grammyDf %>%
  select(Album, Artist, GrammyAward) %>%
  group_by(Artist, Album) %>%
  mutate(numGrammys = n()) %>% 
  select(Album, Artist, numGrammys) 
```
```{r}
grammyDf = grammyDf %>%
  merge(albumYear, by = c("Artist", "Album"))
```



```{r}
grammyDf = grammyDf %>%
  distinct()

accoladesDf = merge(riaaDf , grammyDf, by = c("Artist", "Album"), x.all = TRUE) %>% 
  distinct()

```
```{r}
accoladesDf
```


```{r}
head(songAttrsDf, 90)

attrsOfAlbs = songAttrsDf %>%
  select(-X) %>%
  group_by(Artist, Album) %>%
  summarize(numSongsOnAlb = n(), 
            avgAcousticness = mean(Acousticness), 
            maxAcousticnessTrack = max(Acousticness),
            avgDanceability = mean(Danceability),
            maxDanceabilityTrack = max(Danceability), 
            avgEnergy = mean(Energy), 
            rangeEnergy = max(Energy) - min(Energy), 
            avgInstrumentalness = mean(Instrumentalness),
            maxInstrumentalnessTrack = max(Instrumentalness), 
            avgLiveness = mean(Liveness), 
            maxLivenessTrack = max(Liveness),
            avgLoudness = mean(Loudness),
            rangeLoudness = max(Loudness) - min(Loudness),
            avgSpeechiness = mean(Speechiness), 
            maxSpeechinessTrack = max(Speechiness),
            avgValence = mean(Valence), 
            rangeValence = max(Valence) - min(Valence), 
            totalDurationSeconds = sum(Duration)/1000, 
            avgTempo = mean(Tempo), 
            rangeTempo = max(Tempo) - min(Tempo), 
            moreCommonMajMin = names(which(table(Mode) == max(table(Mode)))[1]), 
            presenceUncommonTS = ifelse(any(TimeSignature != 4), 0, 1), 
            totalExplicit = sum(Explicit, na.rm = TRUE))%>%
  mutate(proportionExplicit =totalExplicit/numSongsOnAlb) %>% 
  merge(albumDates, by = c("Artist", "Album")) %>%
  mutate(ReleaseYear = format(ReleaseDate, "%Y")) %>%
  select(-ReleaseDate) %>%
  merge(yearlyPopularGenre, by.x = "ReleaseYear", by.y = "Year") %>% 
  merge(artistDf %>% select(-X), by = "Artist") 

attrsOfAlbs %>% filter(Artist == "Eminem")
```


Get proportion of songs on each album which were on the hot 100 at some point
```{r}
charted = billboardDf %>% 
  select("BillboardArtist"=Artists,"BillboardSong"= Name) %>% 
  distinct()

artists = strsplit(charted$BillboardArtist, split = ",")
charted = data.frame(BillboardArtist = unlist(artists), BillboardSong = rep(charted$BillboardSong,sapply(artists, length)))

temp = songAttrsDf %>%
  select(Artist, Album, Name) %>%
  group_by(Artist, Album) %>% 
  merge(charted, by.x = c("Artist","Name"), by.y = c("BillboardArtist", "BillboardSong") ) %>%
  group_by(Artist, Album) %>%
  summarize(numChartingSongs = n())

attrsOfAlbsNext =merge(attrsOfAlbs, temp, by = c("Artist", "Album"), all.x = TRUE)
attrsOfAlbsNext[["numChartingSongs"]][is.na(attrsOfAlbsNext[["numChartingSongs"]])] = 0

attrsOfAlbsPropotions = attrsOfAlbsNext %>% 
  group_by(Artist, Album) %>%
  mutate(proportionAlbumCharting = numChartingSongs / numSongsOnAlb)

head(attrsOfAlbsPropotions, 100)
```

```{r}
attrsOfAlbsFinal = attrsOfAlbs %>%
  merge(accoladesDf, by = c("Artist", "Album"), all.x = TRUE)
attrsOfAlbsFinal$numGrammys[is.na(attrsOfAlbsFinal$numGrammys)] = 0

```


```{r}
write_csv(attrsOfAlbsFinal, "Data/attrsOfAlbsFinal.csv")
write_csv(artistDf, "Data/artistDf.csv")
```

- PCA 
- KMeans